Nowcasting Covid-19 statistics reported with delay: a case-study of Sweden

### 2020-06-06 output/paper/paper.tex
Updates from Overleaf

The death counts suffer from the longest reporting day. In their daily presentation, the Swedish Public Health Agency warns for this by stopping the 7-day moving average trend line 10 days before the latest date. But not only are deaths often reported far further back than 10 days, a bar plot still shows the latest information from the most recent days creating a sense of a downward trend. In fact, this might be the reason why the number of deaths has been underestimated repeatedly. At the peak, deaths were initially believed to level out at around 60 per day, but after all cases had been reported more than two weeks later, the actual level was closer to 100 (\cite{Ohman2020_antalet_virusdoda}).

@@

The removal method that has three major advantages over simply reporting moving averages:
\begin{itemize}
	\item it does not relay previous trend in the data,
	\item we can generate confidence bounds for what is the reasonable range of the uncertainty in the event frequency at a given day,
	\item the uncertainty in the estimate can be carried over to epidemiological models that uses the estimate as input, and hence give more realistic models.
\end{itemize}

A classic example where the method proposed to solve this problem has been used is in estimating statistics of trapping a closed population of animals (\cite{Pollock1991_review_papers}). Each day the trapped animals are collected, and kept, and if there is no immigration the number of trapped animals the following days will, on average, decline. This pattern of declining number of trapped animals allows one to draw inference of the underlying population size. Here we replace the animal population with the true number of deaths or cases in a given day. Instead of traps we have the new reports of COVID--19 events. As the number of new reported deaths for a given day declines, we can draw inference on how many actually died that day. In fact, if we assume that the reporting structure is constant over time we can after a while quickly get good estimate of the actual number.

@@

In Figure 1 we apply the model to the latest statistics from Sweden. The graph shows reported and predicted deaths (with uncertainty intervals) as bars and a dashed line plots the 7-day (centered) moving average.

In Figure 1 we illustrate the similarity and difference between the 7-day moving average and the new Bayesian prediction model with 95\% prediction intervals to the reported number of deaths from COVID-19 in Sweden. The model provides estimates of actual deaths considerably above the reported number of deaths and the uncertainty of the estimate.

\begin{figure}
    \centering
    \includegraphics{../plots/}
    \caption{Caption}
    \label{fig:my_label}
\end{figure}

% TODO: FIGURE 1 HERE
%Figure 1. The 7-day moving average (blue) and the prediction from the Bayesian model (black) with prediction intervals overlaid. The grey bars represent the number of reported deaths up to May the 12th, 2020, in COVID-19 in Sweden.

%In Figure 2 we use the method to predict the number of new deaths that will be reported in the most recent report and how their date of death will be distributed over time. It is clear from the model prediction that the delay is substantial and therefore interpretations of the daily reported number of deaths without these models to adjust it are difficult.

% TODO: FIGURE 2 here
%Figure 2. The bars represent deaths in COVID-19 in Sweden and the prediction intervals represent the Bayesian predictions of the number of deaths for the specific day in the latest report.

@@

Nevertheless, the method also has its limitations. As presented, the model assumes that all deaths are reported in the same manner, given there exists many regions in Sweden this is unlikely to be the case. For example, it is easy to see that the Swedish region Västra Götland follows a different reporting structure compared to Stockholm. Building a model for each region separately would most likely give better results and make the assumptions more reasonable. Unfortunately we do not currently have access to the high resolution data required to do so. Another limitation is that the model assumes that the number of new reported deaths for a given day cannot be negative, which is not actually true, due to miscount or mis classification of days. The number of such cases is very small, however, and its removal should not make much difference. The central assumption of the model is that the proportions deaths reported each day is fixed (up to the known covariates). If actual reporting standards change over time and its not explicitly modelled by a covariate, the model will not be able to account for this. But reporting likely becomes faster as the crisis infrastructure improves. One can imagine that after a while the reporting improves, or is changed, if this is not accounted for by a covariate in the model, it will report incorrect numbers. Of course, there might be unknown variables that we have failed to incorporate, but at the least the model is an improvement from the estimates using moving averages.

>>>>

The death counts suffer from the longest reporting delay. In their daily presentation, the Swedish Public Health Agency warns for this by stopping the 7-day moving average trend line 10 days before the latest date. But not only are deaths often reported far further back than 10 days, a bar plot still shows the latest information from the most recent days creating a sense of a downward trend. In fact, this might be the reason why the number of deaths has been underestimated repeatedly. At the peak, deaths were initially believed to level out at around 60 per day, but after all cases had been reported more than two weeks later, the actual level was closer to 100 (\cite{Ohman2020_antalet_virusdoda}).

@@

The removal method that has three major advantages over simply reporting moving averages:

\begin{itemize}
	\item it does not relay previous trend in the data;
	\item we can generate confidence bounds for what is the reasonable range of the uncertainty in the event frequency at a given day;
	\item the uncertainty in the estimate can be carried over to epidemiological models that uses the estimate as input, and hence give more realistic models.
\end{itemize}

A classic example where the method proposed to solve this problem has been used is in estimating statistics of trapping a closed population of animals (\cite{Pollock1991_review_papers}). Each day the trapped animals are collected, and kept, and if there is no immigration the number of trapped animals the following days will, on average, decline. This pattern of declining number of trapped animals allows one to draw inference of the underlying population size. Here we replace the animal population with the true number of deaths, or cases, on a given day. Instead of traps we have the new reports of COVID--19 events. As the number of new reported deaths for a given day declines, we can draw inference on how many actually died that day. In fact, if we assume that the reporting structure is constant over time we can after a while quickly get good estimate of the actual number.

@@

In Figure~\ref{fig:latest_prediction} we apply the model to the latest statistics from Sweden. The graph shows reported and predicted deaths (with uncertainty intervals) as bars and a dashed line plots the 7-day (centered) moving average. A version without predictions is used in the Public Health Agency´s daily press briefings. As expected, the model provides estimates of actual deaths considerably above the reported number of deaths.

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{../plots/latest_prediction}
    \caption{Swedish Covid-19 deaths as of 2020-05-22 and model predictions}
    \label{fig:latest_prediction}
\end{figure}

\chapter{Model Performance}

Figure~\ref{fig:four_dates} depicts four randomly chosen dates where the model is compared to a benchmark prediction based on historical averages. The model and the benchmark are tasked with predicting the total number of individuals who have died at a given date and have been reported within 14 days of that date. As time progresses, more deaths are reported and the dashed grey line approaches the horizontal line. Meanwhile model uncertainty decreases.

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{../plots/lag_prediction_by_date}
    \caption{Model accuracy over time for four dates, compared to benchmark}
    \label{fig:four_dates}
\end{figure}

Figure~\ref{fig:model_metrics} shows model performance compared to the benchmark for three difference performance metrics. All three graphs are based on predictions of reported deaths within 14 days, and show how performance increases as more data has been reported. Each data point is the average of all dates where predictions can be evaluated. SCRPS is a measure of accuracy that rewards precision, similar to a Brier score (see definition in Appendix). The central plot shows the width of the confidence intervals, and the rightmost one the proportion of CI´s that cover the true value.

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{../plots/model_metrics}
    \caption{Model accuracy over time for four dates, compared to a benchmark}
    \label{fig:model_metrics}
\end{figure}

@@

Nevertheless, the method also has its limitations. As presented, the model assumes that all deaths are reported in the same manner, given there exists many regions in Sweden this is unlikely to be the case. For example, it is easy to see that the Swedish region Västra Götland follows a different reporting structure compared to Stockholm. Building a model for each region separately would most likely give better results and make the assumptions more reasonable. Unfortunately we do not currently have access to the high resolution data required to do so. Another limitation is that the model assumes that the number of new reported deaths for a given day cannot be negative, which is not actually true, due to miscount or misclassification of days. The number of such cases is very small, however, and its removal should not make much difference. The central assumption of the model is that the proportions deaths reported each day is fixed (up to the known covariates). If actual reporting standards change over time and its not explicitly modelled by a covariate, the model will not be able to account for this. But reporting likely becomes faster as the crisis infrastructure improves. One can imagine that after a while the reporting improves, or is changed, if this is not accounted for by a covariate in the model, it will report incorrect numbers. Of course, there might be unknown variables that we have failed to incorporate, but at the least the model is an improvement from the estimates using moving averages. When the covariates to the reporting delay pattern is known, the model can incorporate them and provide a more accurate nowcasting of COVID-19 events.
